---
title: "Life History Relationship"
author: "Laurence Kell"
date: "August 13th, 2014"
output: rmarkdown::tufte_handout
---

<!--
%% \VignetteEngine{knitr::knitr}
-->


```{r knitr_init, echo=FALSE, results="asis"}
library(knitr)

# output: rmarkdown::tufte_handout

## Global options
opts_chunk$set(echo    =TRUE,
               eval    =TRUE,
               cache   =FALSE,
               cache.path="cache/",
               prompt  =FALSE,
               comment =NA,
               message =FALSE,
               tidy    =FALSE,
               warnings=FALSE,
               fig.height=4.5,
               fig.width =4.5)
```

# Introduction

Many studies have shown the relationships between life history traits for processes such as growth, maturity and natural mortality. Life history has been used to develop priors in stock assesments and to parameterise ecological models. 

`FLife` has a variety of methods for modelling life history traits and processes and can be used to create `FLBRP` and `FLStock` objects. The `lhPar` method uses life history relationships to fill in missing parameters There are also a variety of methods to model age based processes and to estimate population parameters and reference points. 

```{r}
library(FLCore)
library(FLBRP)
library(FLife)
library(FLBRP)
```

Biological processes can be modelled using functions for growth, maturity and natural mortality, e.g. `vonB`, `sigmoid`, and `lorezen` methods. These take as their first argument an object for age, length or weight and an `FLPar` as the second argument with the life history parameters. 

# Life history relationships

The `lhPar` method uses life history relationships to derive the missing parameters, e.g. if only $L_{\infty}$ is known then parameters for growth, maturity, natural mortality can derived. The first argument to `lhPar` is `param`, an `FLPar` object with as a minimum `linf` of the von Bertalanffy equation, then `k` and length and age at maturity $l50$ $a50$ are derived i.e. 

\begin{equation} k=3.15linf^{-0.64} \end{equation}
 
 Gislason et al 2008
 
 \begin{equation} l_{50}=0.72l_{\infty}^{0.93}\end{equation}
    
 \begin{equation}   a_{50}=log(1-(l50/l_{\infty}/-k+t0) \end{equation}

  
Natural mortality parameters from Model 2, Table 1 Gislason 2010

$m1= 0.55(l_{\infty}^{1.44})k$ and $m2=-1.61$        

Other values are not derived from life history theory, if not supplied by the user, are `a` coefficient of length weight relationship, `b` exponent of length weight relationship, `ato95` age at which 95\% of fish are mature, offset to age at which 50\% are mature, `sl` selectivity-at-age parameter, standard deviation of lefthand limb of double normal, `sr` stock recruitment relationship, `s` steepness of stock recruitment relationship, `v` virgin biomass.
 

```{r lhPar}
par=lhPar(FLPar("linf"=100))
par
```

The `FLPar` can then be used to create vectors for biological processes 

```{r age}
age=FLQuant(0:20,dimnames=list(age=0:20))

ln =vonB(age,par)
mat=sigmoid(age,par)
wt =par["a"]*ln^par["b"]
wt =len2wt(ln,par)
```

```{r, fig.margin=TRUE, fig.cap="Biological properties",echo=FALSE}
ggplot(FLQuants(Length=ln,Mass=wt,Maturity=mat))+
  geom_line(aes(age,data))+
  facet_grid(qname~.,scale="free")+
  xlab("Age")+ylab("")+theme_bw()
```

Natural mortality can either be a function of mass (`lorenzen`) or length (`gislason`)
```{r age-m}
m =FLQuants(Lorenzen=lorenzen(wt,par[c("m1","m2")]),
            Gislason=gislason(ln,par[-(8:9)]))
```

```{r, fig.margin=TRUE, fig.height=4, fig.cap="Natural Mortality",echo=FALSE,eval=FALSE}
ggplot(m)+
  geom_line(aes(age,data,col=qname))+
  xlab("Age")+ylab("Natural Mortality")+theme_bw()+
  theme(legend.title=element_blank(),legend.position="bottom")
```

Selection pattern can be modelled as flat topped or dome shaped as a double normal (´dnormal´)
```{r age-sel}
sel=FLQuants("Flat"=dnormal(age,FLPar(a1="4",sl=3,sr=500)),
             "Dome"=dnormal(age,FLPar(a1="4",sl=3,sr=5)))
```

```{r, fig.margin=TRUE, fig.height=4, fig.cap="Selection patterns",echo=FALSE,eval=FALSE}
ggplot(sel)+
  geom_line(aes(age,data,col=qname))+
  xlab("Age")+ylab("")+theme_bw()
```

Selectivity by default is set so age at peak selectivity is the same as age at 50/% mature (a50) As a minimum all `lhPar` requires is `linf` the asymptotic length of the von Bertalannfy growth equation. 

#FLBRP
```{r}
eql=lhEql(par)
```

#FLStock
```{r om}
library(FLash)
om=fwd(eql)
```

#Leslie Matrix

#Indices
```{r}
fcrash=FLQuant(refpts(eql)["crash","harvest",drop=T],
               dimnames=list(iter=seq( dims(eql)$iter)))

rate =r(leslie(eql,fbar=fcrash))
lop  =FLife:::lopt(par)
```  
  
#Catch curve analysis

```{r,eval=FALSE}
data(bonLn)
rslt=with(subset(bonLn,year==2013), powh(len,n))
```

