---
title: "Life History Relationship"
author: "Laurence Kell"
date: "August 13th, 2014"
output: rmarkdown::tufte_handout
---

<!--
%% \VignetteEngine{knitr::knitr}
-->


```{r knitr_init, echo=FALSE, results="asis"}
library(knitr)

# output: rmarkdown::tufte_handout

## Global options
opts_chunk$set(echo    =TRUE,
               eval    =TRUE,
               cache   =FALSE,
               cache.path="cache/",
               prompt  =FALSE,
               comment =NA,
               message =FALSE,
               tidy    =FALSE,
               warnings=FALSE,
               fig.height=6,
               fig.width =6)
```


```{r,echo=TRUE,eval=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
library(FLCore)
library(FLife)
library(ggplotFL)
```

# Introduction

The importance of life history traits has long been recognied (Holt 1962; Beverton 1963; Cushing 1971) and many studies have shown the relationships between processes such as growth, maturity and natural mortality (e.g. ). These relationships have been used to devlop priors in stock assesments and to parameterise ecological models.

code{FLife} has a variety of methods for modelling life history traits and processes such as growth, maturity and fishery selectivity-at-age. The                                                                                                                                        code{gislasim} method takes an code{FLPar} object with life history parameters and uses relationships between them to fill in missing parameters. A code{FLBRP} can then be generated using code{lh}. 


# Age based processes

```{r age-grow}
data(pars)

age=FLQuant(0:20,dimnames=list(age=0:20))

ln =vonB(age,pars[[1]])

wt=pars[[1]]["a"]*ln^pars[[1]]["b"]
wt=len2wt(ln,pars[[1]])
```

```{r age-mat}
mat=FLife:::sigmoid(age,pars[[1]])
```

```{r}
ggplot(FLQuants(Length=ln,Mass=wt,Maturity=mat))+
  geom_line(aes(age,data))+
  facet_grid(qname~.,scale="free")
```

```{r age-m}
m =FLQuants(Lorenzen=lorenzen(wt,pars[[1]][c("m1","m2")]),
            Gislason=gislasonM(ln,pars[[1]][-(8:9)]))

ggplot(m)+
  geom_line(aes(age,data,col=qname))
```


```{r age-sel}
sel=FLQuants("Flat"=dnormal(age,FLPar(a1="4",sl=3,sr=500)),
             "Dome"=dnormal(age,FLPar(a1="4",sl=3,sr=5)))

ggplot(sel)+
  geom_line(aes(age,data,col=qname))
```

```{r}
dat=data.frame(expand.grid(
                length=seq(10,290,10),year=rep(2001:2010,2)),
                data  =runif(580,0,100)*sample(c(rep(NA,10),rep(10,90)),580,replace=T))
dat=dat[sample(seq(dim(dat)[1],dim(dat)[1])),]

as.FLQuant(dat[sample(seq(dim(dat)[1],dim(dat)[1])),],quant="length")
```

# Life history relationships

code{gislasim} takes an code{FLPar} object with a minimum of code{linf}, it then uses life history relationships to derive the missing parameters for k, age at maturity and M. 

t0=-0.1,a=0.001,b=3,ato95=1,sl=2,sr=5000,s=0.9,v=1000

growth parameters
k=3.15*linf^(-0.64)
From Gislason et al 2008, all species combined
  
Natural mortality parameters from Model 2, Table 1 Gislason 2010
m1= 0.55*(linf^1.44)%*%k
m2=-1.61        

 
maturity parameters from http://www.fishbase.org/manual/FishbaseThe_MATURITY_Table.htm
  if (!("asym"    %in% dimnames(par)$params)) par=rbind(par,FLPar("asym"    =asym, iter=dims(par)$iter))

      $l_{50}=0.72linf^{0.93}$
    
    $a50=log(1-(l50/linf/-k+t0$

selectivity guestimate
a1=par["a50"]
 
attributes(par)$units=c("cm","kg","1000s")
  

Uses life history theory to derive parameters for biological relationships,
 i.e. growth, maturity, natural mortality from.
 

 @param   code{par} code{FLPar} object with parameters for life history equations and selection pattern.
 Need L_infty to estimate other parameters, if any other parameters supplied in code{code} then
 these are not provided by the algorithm 
 @param   code{t0} of von Bertalanffy. This is a default that isnt normally derived
 from life history theory, as are the following args.
 @param   code{a} coefficient of length weight relationship
 @param   code{b} exponent of length weight relationship
 @param   code{ato95} age at which 95\% of fish are mature, offset to age at which 50\% are mature
 @param   code{sl} selectivity-at-age parameter, standard deviation of lefthand limb of double normal
 @param   code{sr} stock recruitment relationship
 @param   code{s} steepness of stock recruitment relationship
 @param   code{v} virgin biomass
 


lh=function(par,
            growth       =vonB,
            fnM         =function(par,len) par["m1"]%*%(exp(log(len)%*%par["m2"])), 
            fnMat        =function(params,data) {
              res=FLQuant(0.5,dimnames=dimnames(data))
              res[data> a50]=1
              res[data< a50]=0
              res},
            fnSel        =dnormal,
            sr           ="bevholt",
            range        =c(min=1,max=40,minfbar=1,maxfbar=40,plusgroup=40),
            spwn         = 0,
            fish         = 0.5, # proportion of year when fishing happens
            units=if("units" %in% names(attributes(par))) attributes(par)$units else NULL,
            ...){
  
  args<-list(...)

  if (("m.spwn" %in% names(args)))
    m.spwn =args[["m.spwn"]]
  else
    m.spwn=FLQuant(spwn, dimnames=list(age=range["min"]:range["max"]))
  
  if (("harvest.spwn" %in% names(args)))
    harvest.spwn =args[["harvest.spwn"]]
  else
    harvest.spwn=FLQuant(spwn, dimnames=list(age=range["min"]:range["max"]))
  
  age=FLQuant(range["min"]:range["max"],
              dimnames=list(age =range["min"]:range["max"],
                            iter=dimnames(par)$iter))
  # Get the lengths through different times of the year
  stocklen   <- growth(par,age+m.spwn) # stocklen is length at spawning time
  catchlen   <- growth(par, age+fish) # catchlen is length when fishing happens
  
  midyearlen <- growth(par, age+0.5) # midyear length used for natural mortality
  
  # Corresponding weights
  swt=exp(log(stocklen%*%par["a"]))%*%par["b"]
  cwt=exp(log(catchlen%*%par["a"]))%*%par["b"]
  if ("bg" %in% dimnames(par)$param)  
    swt=exp(log(stocklen%*%par["a"]))%*%par["bg"]
  warning("FLPar%*%FLQuant operator sets 1st dim name to quant regardless")

  if ("numeric" %in% is(fnM)) m.=FLQuant(fnM,dimnames=dimnames(age)) else{
    if ("len" %in% names(formals(fnM)))   
      m.   =fnM(par=par,len=midyearlen) # natural mortality is always based on mid year length
    else if ("age" %in% names(formals(fnM))){ 
      m.   =fnM(age=age+0.5,par=par) # natural mortality is always based on mid year length
    }else if ("wt" %in% names(formals(fnM)))
      m.   =fnM(par[c("m1","m2")],swt) 
  
  names(dimnames(m.))[1]="age"}

  #age<<-age
#mspwn<<-m.spwn
#return()
  mat. =fnMat(par,age + m.spwn) # maturity is biological therefore + m.spwn 
  sel. =fnSel(par,age + fish) # selectivty is fishery  based therefore + fish

  ## create a FLBRP object to   calculate expected equilibrium values and ref pts
  dms=dimnames(m.)

  res=FLBRP(stock.wt       =swt,
            landings.wt    =cwt,
            discards.wt    =cwt,
            bycatch.wt     =cwt,
            m              =m.,
            mat            =FLQuant(mat., dimnames=dimnames(m.)),
            landings.sel   =FLQuant(sel., dimnames=dimnames(m.)),
            discards.sel   =FLQuant(0,    dimnames=dimnames(m.)),
            bycatch.harvest=FLQuant(0,    dimnames=dimnames(m.)),
            harvest.spwn   =FLQuant(harvest.spwn,    dimnames=dimnames(m.)),
            m.spwn         =FLQuant(m.spwn,    dimnames=dimnames(m.)),
            availability   =FLQuant(1,    dimnames=dimnames(m.)),
            range          =range)
  ## FApex
  #if (!("range" %in% names(args))) range(res,c("minfbar","maxfbar"))[]<-as.numeric(dimnames(landings.sel(res)[landings.sel(res)==max(landings.sel(res))][1])$age)

  ## replace any slot passed in as an arg
  for (slt in names(args)[names(args) %in% names(getSlots("FLBRP"))[names(getSlots("FLBRP"))!="fbar"]])
    slot(res, slt)<-args[[slt]]

  params(res)=propagate(params(res),dims(res)$iter)
Stock recruitment relationship
  model(res) =do.call(sr,list())$model
  
  if (sr=="shepherd" & !("c" %in% names(par))){
    
    dmns=dimnames(par)
    
    dmns$params=c(dmns$params,"c")
    
    par.=FLPar(NA,dimnamels=dmns)
    par.[dimnames(par)$params]=par
    par.["c"]=1
    par=par.}
    
    params(res)=FLPar(c(a=as.numeric(NA),b=as.numeric(NA)),iter=dims(par)$iter)
    for (i in seq(dims(par)$iter))
      if (sr=="shepherd")
        params(res)[,i][]=unlist(c(FLCore:::ab(par[c("s","v","c"),i],sr,spr0=FLCore:::iter(spr0(res),i))[c("a","b","c")]))
    else
      params(res)[,i][]=unlist(c(FLCore:::ab(par[c("s","v"),i],sr,spr0=FLCore:::iter(spr0(res),i))[c("a","b")]))

    warning("iter(params(res),i)=ab(par[c(s,v),i],sr,spr0=iter(spr0(res),i))[c(a,b)] assignment doesnt work")
    warning("iter(FLBRP,i) doesn't work")
  }else{
    if (sr=="shepherd")
      params(res)=FLCore:::ab(par[c("s","v","c")],sr,spr0=spr0(res))[c("a","b","c")]
    else{ 
      params(res)=FLCore:::ab(par[c("s","v")],sr,spr0=spr0(res))[c("a","b")]
      }
    }

  refpts(res)=propagate(FLBRP:::refpts(res)[c("virgin","msy","crash","f0.1","fmax")],dims(par)$iter)
  res=brp(res)

  if ("fbar" %in% names(args)) 
    fbar(res)<-args[["fbar"]] else 
      if (any((!is.nan(FLBRP:::refpts(res)["crash","harvest"])))) 
        fbar(res)<-FLQuant(seq(0,1,length.out=101),quant="age")%*%FLBRP:::refpts(res)["crash","harvest"]
  
  names(dimnames(fbar(res)))[1]="age"
  res=brp(res)
   
  if (!("units" %in% names(attributes(par))))  return(res)
  if (all(is.na(attributes(par)$units)))  return(res)
  
  try(res <- setUnits(res, par),silent=TRUE)
  
  return(res)}
 

```{r lh}

```

# Equilibrium
```{r gisalsim}
```

# Operating Models
```{r om}
```

# Leslie Matrix
```{r leslie}
```

# Reference points and stock indicators
```{r refpt}
```

# Catch curve analyses
```{r cc}
```

# Elasticity Analysis
```{r elas}
```


# Indicators

The length where the biomass of an unfished cohort reaches a maximum is known as $L_{opt}$ is (Beverton and Holt, 1993). It is a reference point for growth overfishing, since if most fish caught are below this length then yield from the fishery is potentially being lost. Depending on the level of infomation $L_{opt}$ can be estimated as   
  
two thirds of $L_{max}$  
  
$L_{\infty}\frac{3}{3+M/k}$   
    
or from a yield per recruit analysis
dimnames(pars[[1]])$params
Where $L_{max}$ is the maximum observed size, $M$ the instantaneous natural mortality and $k$ the growth rate from the Von Bertalannfy growth curve (Von Bertalannfy, 1957).


```{r par,echo=TRUE,eval=!TRUE,message=FALSE,warning=FALSE}
#data(lnMn)
#data(pars)

data(ln)
data(pars)

lines=ldply(pars,function(x) data.frame(Linf =c(x["linf"]),
                                        L50  =c(x["l50"]),
                                        Lopt =c(x["linf"]*2/3),
                                        Lopt2=c(lopt2(x,.6))))
names(lines)[1]="stock"
dat=melt(lines,id="stock")
names(dat)[2]="Quantity"

ggplot(ln)+
  geom_hline(aes(yintercept=value,col=Quantity),
             data=dat,show_guide=T)+
  geom_ribbon(aes(year,ymin=p25,ymax=p75),alpha=.2)+
  geom_line(  aes(year,p50))+
  facet_grid(stock~.,scale="free")+
  theme_bw()+
  xlab("Year")+ylab("Length (cm)")+
  expand_limits(y=0) 
```

# Population Growth Rate



# References

R.~Beverton and S.~Holt.
\newblock \emph{On the dynamics of exploited fish populations}, volume~11.
\newblock Springer, 1993.

Von~Bertalanffy, L. (1957).
\newblock Quantitative laws in metabolism and growth.
\newblock {\em Quarterly Review of Biology}, pages 217--231.

23 Leslie PH (1945) On the use of matrices in certain population mathematics. Biometrika 33: 183–212. doi: 10.1093/biomet/33.3.183 

[9] McAllister MK, Pikitch EK, Babcock EA (2001) Using demographic methods to construct Bayesian priors for the intrinsic rate of increase in the Schaefer model and implications for stock rebuilding. Canadian Journal of Fisheries and Aquatic Sciences 58: 1871–1890. doi: 10.1139/cjfas-58-9-1871 

[39] Picard N, Chagneau P, Mortier F, Bar-Hen A (2009) Finding confidence limits on population growth rates: Bootstrap and analytic methods. Mathematical Biosciences 219: 23–31 doi:10.1016/j.mbs.2009.02.002. doi: 10.1016/j.mbs.2009.02.002 
 