---
title: "Life History Relationship"
author: "Laurence Kell"
date: "August 13th, 2014"
output: rmarkdown::tufte_handout
---

<!--
%% \VignetteEngine{knitr::knitr}
-->


```{r knitr_init, echo=FALSE, results="asis"}
library(knitr)

# output: rmarkdown::tufte_handout

## Global options
opts_chunk$set(echo    =TRUE,
               eval    =TRUE,
               cache   =FALSE,
               cache.path="cache/",
               prompt  =FALSE,
               comment =NA,
               message =FALSE,
               tidy    =FALSE,
               warnings=FALSE,
               fig.height=6,
               fig.width =6)
```


```{r,echo=TRUE,eval=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
library(FLCore)
library(FLife)
library(ggplotFL)
```

# Introduction

The importance of life history traits has long been recognied (Holt 1962; Beverton 1963; Cushing 1971) and many studies have shown the relationships between processes such as growth, maturity and natural mortality (e.g. ). These relationships have been used to devlop priors in stock assesments and to parameterise ecological models.

code{FLife} has a variety of methods for modelling life history traits and processes such as growth, maturity and fishery selectivity-at-age. The                                                                                                                                        code{gislasim} method takes an code{FLPar} object with life history parameters and uses relationships between them to fill in missing parameters. A code{FLBRP} can then be generated using code{lh}. 


# Age based processes

```{r age-grow}
data(pars)

age=FLQuant(0:20,dimnames=list(age=0:20))

ln =vonB(age,pars[[1]])

wt=pars[[1]]["a"]*ln^pars[[1]]["b"]
wt=len2wt(ln,pars[[1]])
```

```{r age-mat}
mat=FLife:::sigmoid(age,pars[[1]])
```

```{r, fig.margin=TRUE, fig.width=3, fig.height=6, fig.cap="Biological properties"}
ggplot(FLQuants(Length=ln,Mass=wt,Maturity=mat))+
  geom_line(aes(age,data))+
  facet_grid(qname~.,scale="free")
```

```{r age-m,eval=FALSE}
m =FLQuants(Lorenzen=lorenzen(wt,pars[[1]][c("m1","m2")]),
            Gislason=gislason(ln,pars[[1]][-(8:9)]))

ggplot(m)+
  geom_line(aes(age,data,col=qname))
```


```{r age-sel}
sel=FLQuants("Flat"=dnormal(age,FLPar(a1="4",sl=3,sr=500)),
             "Dome"=dnormal(age,FLPar(a1="4",sl=3,sr=5)))

ggplot(sel)+
  geom_line(aes(age,data,col=qname))
```

```{r}
dat=data.frame(expand.grid(
                length=seq(10,290,10),year=rep(2001:2010,2)),
                data  =runif(580,0,100)*sample(c(rep(NA,10),rep(10,90)),580,replace=T))
dat=dat[sample(seq(dim(dat)[1],dim(dat)[1])),]

as.FLQuant(dat[sample(seq(dim(dat)[1],dim(dat)[1])),],quant="length")
```

# Life history relationships

code{gislasim} takes an code{FLPar} object with a minimum of code{linf}, it then uses life history relationships to derive the missing parameters for k, age at maturity and M. 

t0=-0.1,a=0.001,b=3,ato95=1,sl=2,sr=5000,s=0.9,v=1000

growth parameters
k=3.15*linf^(-0.64)
From Gislason et al 2008, all species combined
  
Natural mortality parameters from Model 2, Table 1 Gislason 2010
m1= 0.55*(linf^1.44)%*%k
m2=-1.61        

 
maturity parameters from http://www.fishbase.org/manual/FishbaseThe_MATURITY_Table.htm
  if (!("asym"    %in% dimnames(par)$params)) par=rbind(par,FLPar("asym"    =asym, iter=dims(par)$iter))

      $l_{50}=0.72linf^{0.93}$
    
    $a50=log(1-(l50/linf/-k+t0$

selectivity guestimate
a1=par["a50"]
 
attributes(par)$units=c("cm","kg","1000s")
  

Uses life history theory to derive parameters for biological relationships,
 i.e. growth, maturity, natural mortality from.
 

 @param   code{par} code{FLPar} object with parameters for life history equations and selection pattern.
 Need L_infty to estimate other parameters, if any other parameters supplied in code{code} then
 these are not provided by the algorithm 
 @param   code{t0} of von Bertalanffy. This is a default that isnt normally derived
 from life history theory, as are the following args.
 @param   code{a} coefficient of length weight relationship
 @param   code{b} exponent of length weight relationship
 @param   code{ato95} age at which 95\% of fish are mature, offset to age at which 50\% are mature
 @param   code{sl} selectivity-at-age parameter, standard deviation of lefthand limb of double normal
 @param   code{sr} stock recruitment relationship
 @param   code{s} steepness of stock recruitment relationship
 @param   code{v} virgin biomass
 


lh=function(par,
            growth       =vonB,
            fnM         =function(par,len) par["m1"]%*%(exp(log(len)%*%par["m2"])), 
            fnMat        =function(params,data) {
              res=FLQuant(0.5,dimnames=dimnames(data))
              res[data> a50]=1
              res[data< a50]=0
              res},
            fnSel        =dnormal,
            sr           ="bevholt",
            range        =c(min=1,max=40,minfbar=1,maxfbar=40,plusgroup=40),
            spwn         = 0,
            fish         = 0.5, # proportion of year when fishing happens
            units=if("units" %in% names(attributes(par))) attributes(par)$units else NULL,
            ...){
  
  args<-list(...)

  if (("m.spwn" %in% names(args)))
    m.spwn =args[["m.spwn"]]
  else
    m.spwn=FLQuant(spwn, dimnames=list(age=range["min"]:range["max"]))
  
  if (("harvest.spwn" %in% names(args)))
    harvest.spwn =args[["harvest.spwn"]]
  else
    harvest.spwn=FLQuant(spwn, dimnames=list(age=range["min"]:range["max"]))
  
  age=FLQuant(range["min"]:range["max"],
              dimnames=list(age =range["min"]:range["max"],
                            iter=dimnames(par)$iter))
  # Get the lengths through different times of the year
  stocklen   <- growth(par,age+m.spwn) # stocklen is length at spawning time
  catchlen   <- growth(par, age+fish) # catchlen is length when fishing happens
  
  midyearlen <- growth(par, age+0.5) # midyear length used for natural mortality
  
  # Corresponding weights
  swt=exp(log(stocklen%*%par["a"]))%*%par["b"]
  cwt=exp(log(catchlen%*%par["a"]))%*%par["b"]
  if ("bg" %in% dimnames(par)$param)  
    swt=exp(log(stocklen%*%par["a"]))%*%par["bg"]
  warning("FLPar%*%FLQuant operator sets 1st dim name to quant regardless")

  if ("numeric" %in% is(fnM)) m.=FLQuant(fnM,dimnames=dimnames(age)) else{
    if ("len" %in% names(formals(fnM)))   
      m.   =fnM(par=par,len=midyearlen) # natural mortality is always based on mid year length
    else if ("age" %in% names(formals(fnM))){ 
      m.   =fnM(age=age+0.5,par=par) # natural mortality is always based on mid year length
    }else if ("wt" %in% names(formals(fnM)))
      m.   =fnM(par[c("m1","m2")],swt) 
  
  names(dimnames(m.))[1]="age"}

  #age<<-age
#mspwn<<-m.spwn
#return()
  mat. =fnMat(par,age + m.spwn) # maturity is biological therefore + m.spwn 
  sel. =fnSel(par,age + fish) # selectivty is fishery  based therefore + fish

  ## create a FLBRP object to   calculate expected equilibrium values and ref pts
  dms=dimnames(m.)

  res=FLBRP(stock.wt       =swt,
            landings.wt    =cwt,
            discards.wt    =cwt,
            bycatch.wt     =cwt,
            m              =m.,
            mat            =FLQuant(mat., dimnames=dimnames(m.)),
            landings.sel   =FLQuant(sel., dimnames=dimnames(m.)),
            discards.sel   =FLQuant(0,    dimnames=dimnames(m.)),
            bycatch.harvest=FLQuant(0,    dimnames=dimnames(m.)),
            harvest.spwn   =FLQuant(harvest.spwn,    dimnames=dimnames(m.)),
            m.spwn         =FLQuant(m.spwn,    dimnames=dimnames(m.)),
            availability   =FLQuant(1,    dimnames=dimnames(m.)),
            range          =range)
  ## FApex
  #if (!("range" %in% names(args))) range(res,c("minfbar","maxfbar"))[]<-as.numeric(dimnames(landings.sel(res)[landings.sel(res)==max(landings.sel(res))][1])$age)

  ## replace any slot passed in as an arg
  for (slt in names(args)[names(args) %in% names(getSlots("FLBRP"))[names(getSlots("FLBRP"))!="fbar"]])
    slot(res, slt)<-args[[slt]]

  params(res)=propagate(params(res),dims(res)$iter)
Stock recruitment relationship
  model(res) =do.call(sr,list())$model
  
  if (sr=="shepherd" & !("c" %in% names(par))){
    
    dmns=dimnames(par)
    
    dmns$params=c(dmns$params,"c")
    
    par.=FLPar(NA,dimnamels=dmns)
    par.[dimnames(par)$params]=par
    par.["c"]=1
    par=par.}
    
    params(res)=FLPar(c(a=as.numeric(NA),b=as.numeric(NA)),iter=dims(par)$iter)
    for (i in seq(dims(par)$iter))
      if (sr=="shepherd")
        params(res)[,i][]=unlist(c(FLCore:::ab(par[c("s","v","c"),i],sr,spr0=FLCore:::iter(spr0(res),i))[c("a","b","c")]))
    else
      params(res)[,i][]=unlist(c(FLCore:::ab(par[c("s","v"),i],sr,spr0=FLCore:::iter(spr0(res),i))[c("a","b")]))

    warning("iter(params(res),i)=ab(par[c(s,v),i],sr,spr0=iter(spr0(res),i))[c(a,b)] assignment doesnt work")
    warning("iter(FLBRP,i) doesn't work")
  }else{
    if (sr=="shepherd")
      params(res)=FLCore:::ab(par[c("s","v","c")],sr,spr0=spr0(res))[c("a","b","c")]
    else{ 
      params(res)=FLCore:::ab(par[c("s","v")],sr,spr0=spr0(res))[c("a","b")]
      }
    }

  refpts(res)=propagate(FLBRP:::refpts(res)[c("virgin","msy","crash","f0.1","fmax")],dims(par)$iter)
  res=brp(res)

  if ("fbar" %in% names(args)) 
    fbar(res)<-args[["fbar"]] else 
      if (any((!is.nan(FLBRP:::refpts(res)["crash","harvest"])))) 
        fbar(res)<-FLQuant(seq(0,1,length.out=101),quant="age")%*%FLBRP:::refpts(res)["crash","harvest"]
  
  names(dimnames(fbar(res)))[1]="age"
  res=brp(res)
   
  if (!("units" %in% names(attributes(par))))  return(res)
  if (all(is.na(attributes(par)$units)))  return(res)
  
  try(res <- setUnits(res, par),silent=TRUE)
  
  return(res)}
 

```{r lh}

```

# Equilibrium
```{r gisalsim}
```

# Operating Models
```{r om}
```

# Leslie Matrix
```{r leslie}
```

The maximum theoretical rate of increase of a population in the absence of density-dependent regulation is given by

\begin{equation} r=\frac{dN}{dt} \frac{1}{N}  \end{equation}

\noindent where the intrinsic population growth rate ($r$) is a function of population size ($N$) and $\frac{dN}{dt}$ the instantaneous rate of increase of the population.

In ecology Leslie matrices \citep{leslie1945use} are widely used to estimate r \citep{picard2009finding} and have been used in fisheries to develop Bayesian priors for r for use in stock assessment citep[see][]{mcallister2001using}. The Leslie Matrix ($A$) is a transition matrix that models age dynamics and can be specified either in numbers or biomass. Each age-class is described by a vector ($B_t$) of length p equal to the terminal age. Entries in the matrix are fecundity ($f_i$) (the quantity of age zero females produced per unit of mature biomass by each age-class) and the survival (and growth if biomass) of an age-class ($s_i$) in each time step $i$, i.e.

\begin{equation}
\begin{array}{l}
b_{1}= f_{2}b_{2}+ ... + f_{3}b_{p}\\
b_{2}=s_{1}b_{1}\\
~~~~~~...\\
b_{p}=s_{p-1}b_{p-1}+s_{p}b_{p}\\
\end{array}
\end{equation}

The matrix of this linear system is 
\begin{equation}
\mathbf{A}=%
\begin{pmatrix}
0 & f_{2} & ... & f_{p} \\ 
s_{1} & 0 & ... & 0 \\ 
 &...&\\
0 & ... & s_{p-1} & s_{p}%
\end{pmatrix}%
\end{equation}

If the initial population is

\begin{equation}
\mathbf{B}^{0}=%
\begin{pmatrix}
b_{1} \\ 
b_{2} \\ 
...\\
b_{p}%
\end{pmatrix}%
\end{equation}

then after time step i=1 the population is given by

\begin{equation}
\mathbf{B}^{i}=\mathbf{A}^{i}\mathbf{B}^{0}
\end{equation}

As $i$ tends to infinity the system reaches equilibrium and the contribution of each age group in the population becomes stable. The population growth rate $r$ is then derived from $\lambda$ the dominant eigenvalue of $A$ \citep{caswell1989matrix}.

To construct the Leslie matrix requires estimates of $f_i$ and $p_i$. In this study these were derived by combining a stock recruitment relationship with a spawner-per-recruit ($S/R$) and yield-per-recruit ($Y/R$) analyses. The life history parameters were used to derive mass ($W$), proportion mature ($Q$), natural mortality ($M$) and  fishing mortality ($F$) at age. 

The second term is the plus-group, i.e. the summation of all ages from the last age to infinity. 

Growth in length is modelled by the Von Bertalanffy growth equation \cite{von1957quantitative}

\begin{equation} L = L_{\infty}(1 - exp(-k(t-t_0)) \end{equation}
         
where $k$ is the rate at which the rate of growth in length declines as length approaches the asymptotic length  $L_{\infty}$ and $t_{0}$ is the hypothetical time at which an individual is of zero length.

Length is converted to mass using the length-weight relationship 
    
\begin{equation} W = aL_t^b \end{equation}

\noindent where $a$ is the condition factor and $b$ is the allometric growth coefficient.

\cite{gislason2010does} showed that M is significantly related to body length, asymptotic length and k. Temperature is non-significant when k is included, since k itself is correlated with temperature. We therefore model M as

\begin{equation}
            M =  0.55L^{1.61}L_{\infty}^{1.44}k
\end{equation} 

Selection pattern of the fishery was represented by a double normal \citep[see][]{Hilbornetal2000}) with three parameters that describe the age at maximum selection ($a1$), the rate at which the left-hand  limb increases ($sl$) and the right-hand limb decreases ($sr$) which allows flat topped or domed shaped selection patterns to be chosen.
         
\begin{equation}
f(x) = \left\{ \begin{array}{ll}
  		0                                 &\mbox{ if $(a_{50}-x)/a_{95} >  5$} \\
			a_{\infty}                        &\mbox{ if $(a_{50}-x)/a_{95} < -5$} \\
			\frac{m_{\infty}}{1.0+19.0^{(a_{50}-x)/_{95})}} &\mbox{ otherwise}
		\end{array}
       \right.
\end{equation}

The relationship between stock and recruitment was modelled by a Beverton and Holt stock-recruitment relationship \citep{beverton1993dynamics} reformulated in terms of steepness ($h$), virgin biomass ($v$) and $S/R_{F=0}$

\begin{equation}
R=\frac{0.8R_0h}{0.2S/R_{F=0}R_0(1-h)+(h-0.2)S}
\end{equation} 

\noindent where steepness is the ratio of recruitment at 20\% of virgin biomass to virgin recruitment ($R_0$) and $S/R_{F=0}$ is the spawner per recruit at virgin biomass, i.e. when fishing mortality is zero. Steepness is difficult to estimate from stock assessment data sets as there is often insufficient contrast in biomass levels required for its estimation \cite{pepin2015reconsidering}.

S is spawning stock biomass, the sum of the products of the numbers of females, $N$, proportion mature-at-age, $Q$ and their mean fecundity-at-age, $F$, i.e. 

\begin{equation} S = \sum\limits_{i=0}^{p} {N_iQ_iF_i} \end{equation}

where fecundity-at-age is assumed proportional to biomass and the sex ratio to be 1:1. Proportion mature is 50\% at the age that attains a length of $l50$, 0\% below this age and 100\% above. 

# Reference points and stock indicators
```{r refpt}
```

# Catch curve analyses
```{r cc}
```

# Elasticity Analysis
```{r elas}
```


# Indicators

The length where the biomass of an unfished cohort reaches a maximum is known as $L_{opt}$ is (Beverton and Holt, 1993). It is a reference point for growth overfishing, since if most fish caught are below this length then yield from the fishery is potentially being lost. Depending on the level of infomation $L_{opt}$ can be estimated as   
  
two thirds of $L_{max}$  
  
$L_{\infty}\frac{3}{3+M/k}$   
    
or from a yield per recruit analysis
dimnames(pars[[1]])$params
Where $L_{max}$ is the maximum observed size, $M$ the instantaneous natural mortality and $k$ the growth rate from the Von Bertalannfy growth curve (Von Bertalannfy, 1957).


```{r par,echo=TRUE,eval=!TRUE,message=FALSE,warning=FALSE}
#data(lnMn)
#data(pars)

data(ln)
data(pars)

lines=ldply(pars,function(x) data.frame(Linf =c(x["linf"]),
                                        L50  =c(x["l50"]),
                                        Lopt =c(x["linf"]*2/3),
                                        Lopt2=c(lopt2(x,.6))))
names(lines)[1]="stock"
dat=melt(lines,id="stock")
names(dat)[2]="Quantity"

ggplot(ln)+
  geom_hline(aes(yintercept=value,col=Quantity),
             data=dat,show_guide=T)+
  geom_ribbon(aes(year,ymin=p25,ymax=p75),alpha=.2)+
  geom_line(  aes(year,p50))+
  facet_grid(stock~.,scale="free")+
  theme_bw()+
  xlab("Year")+ylab("Length (cm)")+
  expand_limits(y=0) 
```

# Population Growth Rate



# References

R.~Beverton and S.~Holt.
\newblock \emph{On the dynamics of exploited fish populations}, volume~11.
\newblock Springer, 1993.

Von~Bertalanffy, L. (1957).
\newblock Quantitative laws in metabolism and growth.
\newblock {\em Quarterly Review of Biology}, pages 217--231.

23 Leslie PH (1945) On the use of matrices in certain population mathematics. Biometrika 33: 183–212. doi: 10.1093/biomet/33.3.183 

[9] McAllister MK, Pikitch EK, Babcock EA (2001) Using demographic methods to construct Bayesian priors for the intrinsic rate of increase in the Schaefer model and implications for stock rebuilding. Canadian Journal of Fisheries and Aquatic Sciences 58: 1871–1890. doi: 10.1139/cjfas-58-9-1871 

[39] Picard N, Chagneau P, Mortier F, Bar-Hen A (2009) Finding confidence limits on population growth rates: Bootstrap and analytic methods. Mathematical Biosciences 219: 23–31 doi:10.1016/j.mbs.2009.02.002. doi: 10.1016/j.mbs.2009.02.002 
 