---
title: "Life History Relationship"
author: "Laurence Kell"
date: "August 13th, 2014"
output: rmarkdown::tufte_handout
bibliography: ref.bib
---

<!--
%% \VignetteEngine{knitr::knitr}
-->

```{r knitr_init, echo=FALSE, results="asis"}
library(knitr)

# output: rmarkdown::tufte_handout

## Global options
opts_chunk$set(echo    =!TRUE,
               eval    =TRUE,
               cache   =!FALSE,
               cache.path="vignette/",
               prompt  =FALSE,
               comment =NA,
               message =FALSE,
               tidy    =FALSE,
               warnings=FALSE,
               fig.height=4.5,
               fig.width =4.5,
               fig.path  ="tex/")
```

# Introduction

`FLife` package brings together a variety of methods for modelling life history traits and ecological processes and can be used to create `FLR` objects such as `FLBRP` and `FLStock` to model specie, population or stock dynamics. 

Many studies have shown relationships between life history traits such as growth, maturity and natural mortality. This knowledge is of use for providing advice for data poor stocks, developing priors for use in data rich stock assesments and to parameterise ecological models. 

```{r,echo=FALSE}
library(FLCore)
library(FLash)
library(FLBRP)
library(FLife)
library(popbio)
```

# Life history relationships

The `lhPar` method uses life history relationships to fill in missing values in data poor datasets, there are methods to life history parameters, to model age based processes, simulate time series and estimate reference points and indicators. 

In data poor situtuations only the maximum size ($l_{max}$) may be known. Life history relationships  allow parameters for growth, maturity, natural mortality to derived from empirical relationship.

[@gislason2008coexistence] derived relationships to estimate $k$ of the [@vonbert1957quantitative] growth equation 

\begin{equation} k=3.15l_{\infty}^{-0.64} \end{equation}

and maturity from 
\begin{equation} l_{50}=0.72l_{\infty}^{0.93} \end{equation}

from $l_{\infty}$, and natural mortality from length ($l$)

\begin{equation} M=0.55(l-1.66l_{\infty}^{1.44}k\end{equation}

where $m_1= 0.55(l_{\infty}^{1.44})k$ and $m_2=-1.61$        

There are alternative relationship, for example [@lorenzen2002density] derived $M$ from mass (w) 
\begin{equation} M=m_1*w^m_2\end{equation}

The `lhPar` method takes as its first argument an `FLPar` object with as a minimum a value for `linf` and uses these relationships to derive `k` and $l50$


```{r lhPar, echo=TRUE}
par=lhPar(FLPar(linf=100))
```

There are defaults for other values, which if not supplied can not be derived from life history theory. These include $a$ and $b$ from the length weight relationship $w=al^b$, `ato95` the age at which 95\% of fish are mature, offset to age at which 50\% are mature, `sl` selectivity-at-age parameter, standard deviation of lefthand limb of double normal, `sr` stock recruitment relationship, `s` steepness of stock recruitment relationship, `v` virgin biomass.

Biological processes can then be modelled using functions for growth, maturity and natural mortality, e.g. `vonB`, `sigmoid`, and `lorezen` methods. These take as their first argument an object for age, length or weight and an `FLPar` as the second argument with the life history parameters. 


```{r age, echo=TRUE}
age=FLQuant(0:20,dimnames=list(age=0:20))

ln =vonB(age,par)
mat=sigmoid(age,par)
wt =par["a"]*ln^par["b"]
wt =len2wt(ln,par)
```


```{r biol, fig.margin=TRUE, fig.cap="Biological properties",echo=FALSE}
ggplot(FLQuants(Length=ln,Mass=wt,Maturity=mat))+
  geom_line(aes(age,data))+
  facet_grid(qname~.,scale="free")+
  xlab("Age")+ylab("")+theme_bw()
```

Natural mortality can either be a function of mass or length 
```{r age-m}
m =FLQuants(Lorenzen=lorenzen(wt,par[c("m1","m2")]),
            Gislason=gislason(ln,par[-(8:9)]))
```

```{r, fig.margin=TRUE, fig.height=4, fig.cap="Natural Mortality",echo=FALSE}
ggplot(m)+
  geom_line(aes(age,data,col=qname))+
  xlab("Age")+ylab("Natural Mortality")+theme_bw()+
  theme(legend.title=element_blank(),legend.position="bottom")
```

and selection pattern can be modelled as flat topped or dome shaped by using the double normal function

```{r age-sel}
sel=FLQuants("Flat"=dnormal(age,FLPar(a1="4",sl=3,sr=500)),
             "Dome"=dnormal(age,FLPar(a1="4",sl=3,sr=5)))
```

```{r, fig.margin=TRUE, fig.height=4, fig.cap="Selection patterns",echo=FALSE,eval=FALSE}
ggplot(sel)+
  geom_line(aes(age,data,col=qname))+
  xlab("Age")+ylab("")+theme_bw()
```



#FLBRP
```{r}
eql=lhEql(par)
```

#FLStock
```{r om}
library(FLash)
om=fwd(eql)
```

#Examples

##Simulation

```{r}
library(FLife)
library(ggplotFL)

par=lhPar(FLPar(linf=100))
eql=lhEql(par)
```


```{r biol2}
ggplot(FLQuants(eql,"m","mat","stock.wt","catch.sel"))+
  geom_line(aes(age,data))+
  facet_wrap(~qname,scale="free")+
  xlab("Age")+ylab("")+theme_bw()
```

```{r brp}
plot(eql)+theme_bw()+theme(legend.position="bottom")
```

```{r,eval=FALSE}
load("/home/laurie/Desktop/flr/git/FLife/data/teleost.RData")

load("/home/laurie/Desktop/MEGA/papers/generic/fishnets-master/data/fishbase-web/fishbase-web.RData")
cod=apply(subset(fb,species=="Gadus morhua")[,c("linf","k","t0","a","b","lmat")],2,function(x) mean(x,na.rm=T))

cod=data.frame(species="Gadus morhua",as.data.frame(t(cod)))
 names(cod)[7]="l50"
cod[,"t0"]=-0.75

teleost=rbind.fill(teleost,cod)
ach=subset(teleost,species%in%c("Clupea harengus harengus","Gadus morhua",
                             "Thunnus alalunga"))
save(ach,file="/home/laurie/Desktop/flr/git/FLife/data/ach.RData")
```

```{r ts}
load("/home/laurie/Desktop/flr/git/FLife/data/ach.RData")

yrs  =FLQuant(0,dimnames=list(year=1:500))
srDev=rlnorm(1,yrs,.3)
fDev =rlnorm(1,yrs,.2)

scen=expand.grid(Species  =c("Clupea harengus harengus","Thunnus alalunga"),
                 Selection=c("Mature","Juvenile"),
                 FMSY     =c(0.1,1,2))
            
nms=c("linf","k","t0","a","b","l50")  

#attach(scen[1,])

sim=function(Species,Selection,FMSY){
  par    =lhPar(ach[ach$species==Species,nms])
  eql    =lhEql(par)
  
  landings.sel(eql)                              =ifelse(Selection=="Mature",1,0.01)
  landings.sel(eql)[ac(0:as.integer(par["a50"]))]=ifelse(Selection=="Mature",0.01,1)
  
  range(eql)[c("minfbar","maxfbar")]=as.integer(par["a50"])+ifelse(Selection=="Mature",1,0)
  
  eql=brp(eql)
  
  if (FMSY<2)
     fbar(eql)=FLQuant(FMSY,dimnames=list(year=0:500))*FLBRP:::refpts(eql)["msy","harvest"]
  else  
     fbar(eql)=FLQuant(0.75,dimnames=list(year=0:500))*FLBRP:::refpts(eql)["crash","harvest"]
  
  eql=brp(eql)
  
  stk=fwd(eql,f=fbar(eql)[,-1])
  stk=fwd(stk,f=fbar(eql)[,-(1:2)]%*%fDev[,-1],sr=eql,sr.residuals=srDev)

  model.frame(FLQuants(rec  =rec(  stk)/refpts(eql)["virgin","rec"],
                       ssb  =ssb(  stk)/refpts(eql)["msy",   "ssb"],
                       stock=stock(stk)/refpts(eql)["msy",   "biomass"],
                       catch=catch(stk)/refpts(eql)["msy",   "yield"],
                       f    =fbar( stk)/refpts(eql)["msy",   "harvest"]),drop=TRUE)}

sims=mdply(scen,sim)


ggplot(ddply(sims,.(Species,FMSY,Selection), transform, data=ssb))+
  geom_line(aes(year,data,col=Species))+
  facet_grid(FMSY~Selection,scale="free")+
  xlab("Year")+ylab("SSB")+
  scale_x_continuous(limits=c(300,350))+
  theme_bw()+theme(legend.position="bottom")
```

```{r}
ggplot(ddply(sims,.(Species,FMSY,Selection), with, FLife:::spectra(ssb)))+
  geom_line(aes(f,mx,col=Species))+
  facet_grid(FMSY~Selection,scale="free")+
  xlab("Frequency")+ylab("SSB")+ theme(legend.position="bottom")+
  scale_x_continuous(limits=c(.0,.05))

```

```{r}
stk=as(eql,"FLStock")
plot(stk)
```


```{r}
library(FLash)
srDev=rlnorm(100,rec(stk)*0,.3)[,-1]
```

```{r}
plot(stk)
```

##Leslie Matrix

##Data poor
```{r,eval=FALSE}
data(bonLn)
rslt=with(subset(bonLn,year==2013), powh(len,n))
```

##Indices
```{r}
fcrash=FLQuant(refpts(eql)["crash","harvest",drop=T],
               dimnames=list(iter=seq( dims(eql)$iter)))

rate =r(leslie(eql,fbar=fcrash))
lop  =FLife:::lopt(par)
```  
  

#Catch curve analysis


# References


# Data

```{r}
library("GGally")
library(ggbiplot)

load("/home/laurie/Desktop/flr/git/FLife/data/teleost.RData")

ggpairs(teleost[c("l50","linf","k","b")],
  lower=list(params=c(colour="blue")),#,continuous="smooth",se=FALSE),
  diag=list(continuous="bar",params=c(colour="blue")), 
  upper=list(params=list(corSize=6)), axisLabels='none')+
  theme(legend.position = "none", 
       panel.grid.major = element_blank(), 
       axis.ticks       = element_blank(), 
       panel.border     = element_rect(linetype = 1, colour="black", fill=NA))+
  theme_bw(14)
```

```{r lhPar2}
teleost=transform(teleost,l50linf=l50/linf)

family=transform(teleost,what=ifelse(family%in%c("Clupeidae","Scombridae"),ac(family),"other"))

pc=princomp(teleost[,c("linf","k","b","l50linf")],
            cor=TRUE,use="pairwise.complete.obs")
gg=ggbiplot(pc, obs.scale=1, var.scale=1, 
                ellipse=TRUE, ellipse.prob=.5, circle=FALSE,
         groups=factor(family[,"what"]))+
  kobe:::theme_ms(10,legend.position="bottom")

#gg$layers[[2]]$mapping$colour=NULL
#gg$layers[[3]]$mapping$colour=NULL
gg+theme(legend.position="none") #guides(col=guide_legend(nrow = 3))
```

```{r,eval=!FALSE}
par=lhPar(teleost[,c("linf","k","t0","l50","a","b")])

attributes(par)$species=teleost[,"species"]
attributes(par)$family =teleost[,"family"]

mGislason=function(length,params) 
   0.55*(length^-1.66)%*%(params["linf"]^1.44)%*%params["k"]
ref=FLife:::lhRef(par,m=mGislason)
```

```{r lhPop}
dat=ref[dimnames(ref[!is.na(ref[,"rc"]),])[[1]],]
           
pc=princomp(dat[,c("r","rc","lopt","sk")],cor=TRUE,use="pairwise.complete.obs")
gg=ggbiplot(pc, obs.scale=1, var.scale=1, 
                ellipse=TRUE, ellipse.prob=.5, circle=FALSE,
         groups=factor(family[,"what"]) )+
  kobe:::theme_ms(12,legend.position="bottom")

#gg$layers[[2]]=NULL
#gg$layers[[2]]$mapping$colour=NULL
#gg$layers[[3]]$mapping$colour=NULL
gg+theme(legend.position="none")
```

#Elasticity

A measure of ???proportional?????? effect, i.e., the effect that a change in a given matrix element has as a proportional to the change in that element

```{r elasticity}
library(ggplot2)
library(FLCore)
library(FLBRP)
library(FLife)
library(numDeriv)
library(popbio)

source('~/Desktop/flr/git/FLife/R/lhEql.R')
    loptAge=function(params,
                   m     =function(length,params) params["m1"]%*%(exp(log(length)%*%params["m2"])),
                   growth=vonB,
                   ...){   

      loptFn=function(x,params,m){
        
        age   =0:ceiling(x)
        dmns  =list(age=age)
        length=vonB(age=FLQuant(pmin(age+0.5,x),dimnames=dmns),params=params)
        m.    =FLQuant(m(length,params),    dimnames=dmns)
        mCum  =FLQuant(aaply(m.,6,sum))
        n     =exp(-mCum)
        c(n*len2wt(length[ac(ceiling(x))],params))}
            
      dmns=dimnames(params)
      dmns$params="lopt"
      dm  =dim(params)
            
      res=aaply(params,seq(length(dm))[-1],function(x){
            x.=FLPar(x)
            rtn=try(optimise(loptFn,c(0,40),params=x.,maximum=TRUE,m=m)$maximum)
            if ("character" %in% mode(rtn)) rtn=NA
            rtn})
            
      vonB(FLQuant(FLPar(array(res,dim=c(1, dm[-1]),dimnames=dmns))),params)
      
      }

par=lhPar(FLPar(linf=100))

fn=function(par,all=FALSE){
  eql=lhEql(par)
  
  ref=refpts(eql)[,c(1:2,4:5)]
  h  =melt(ref[,2]/ref[,4])[,-3]
  h$quantity="harvest"
  
  dimnames(ref)[[2]][1]="f"
  ref=melt(ref[drop=T])
  
  lop=loptAge(par)
  
  f  =refpts(eql)["crash","harvest"]
  r  =log(lambda(leslie(eql,fbar=FLQuant(c(f)))[drop=T]))
  
  f  =refpts(eql)["msy","harvest"]
  rc =log(lambda(leslie(eql,fbar=FLQuant(c(f)))[drop=T]))
  
  res=rbind(ref,h,
      cbind("refpt"="r",   "quantity"="pop rate", "value"=melt(r[drop=T])),
      cbind("refpt"="rc",  "quantity"="pop rate", "value"=melt(rc[drop=T])),
      cbind("refpt"="lopt","quantity"="length",   "value"=melt(lop[drop=T])))
  
  res=rbind(subset(res,refpt%in%c("r","rc","lopt")),
            subset(res,refpt%in%c("msy")&quantity=="ssb"))
  
  res[,3]=log(res[,3])
  
  if(all) res else c(res[,3])}
  
val=fn(par,TRUE)
jcb=jacobian(fn,par)

dimnames(jcb)=list(y=val$ref,x=dimnames(par)$params)
jcb=melt(jcb)
jcb$quantity=val$quantity

#ggplot(subset(jcb,quantity!="pop rate"))+
#  geom_point(aes(x,y,size=log(abs(value))))

#ggplot(subset(jcb,quantity=="pop rate"))+
#  geom_point(aes(x,y,size=log(abs(value))))

ggplot(ddply(jcb,.(y),transform, value=value/max(abs(value))))+
  geom_point(aes(x,y,size=abs(value)))+
  scale_size_area()+
  theme_bw()+
  theme(legend.position="none")

```

# Stability

An important factor determining a population’s response to perturbation is stability which can be measured in a variety of ways (e.g. Pimm 1984).  In its simplest form, a population can be considered stable if it returns to equilibrium after a perturbation.  Other definitions expand on this and involve the time taken to return to equilibrium after a perturbation, known as the characteristic return time or population resilience.  The lower the characteristic return time, or higher the resilience, the more stable the population.  The stability of a population is strongly influenced by the life history of the population and also the pattern of density dependence.  For some population models the stability is a good indicator of a population’s response to noise (Taylor 1992), but generally stability is insufficient on its own to predict the response (Horwood 1993). Here we use it to indicate how quickly management can cause an effect in a population, e.g. to recover a stock to a level that would support MSY.  In this way, stability can be used as a guide to how controllable the stock is.
For discrete, structured populations this can be calculated using the magnitude of the dominant eigenvalue of the Jacobian matrix evaluated at the equilibrium point (Beddington 1974; Caswell 2001).  If the magnitude of this is less than 1 the population will return to equilibrium after a disturbance, with the stability decreasing as the magnitude approaches 1.  When the magnitude of the dominant eigenvalue is 1 there is a bifurcation and past this point non-equilibrium dynamics, including extinction, are seen.  


```{r,init}
library(ggplot2)
library(ggbiplot)
library("GGally")
library(plyr)
library(reshape)
library(FLCore)
library(FLBRP)
library(FLash)
library(FLife)
library(ggplotFL)
library(kobe)
library(popbio)

data(teleost)
teleost=transform(teleost,l50linf=l50/linf)

alb=FLPar(unlist(teleost[teleost$species=="Thunnus alalunga",
        c("linf","k","t0","l50","a","b")]))
alb=lhPar(rbind(alb,FLPar(m1=0.15,m2=-0.288,s=0.75)))

i=0
```


```{r fig1,fig.height=6,fig.width=6}
ggpairs(log(100*teleost[,c("b","l50","linf","k","l50linf")]),
  lower=list(continuous="smooth", alb=c(colour="blue",se=FALSE)),
  diag=list(continuous="bar",alb=c(colour="blue")), 
  upper=list(alb=list(corSize=6)), axisLabels='none')+  
  #scale_x_log10()+scale_y_log10()
  theme(legend.position = "none", 
        panel.grid.major = element_blank(), 
        axis.ticks       = element_blank(), 
        panel.border     = element_rect(linetype = 1, colour="black", fill=NA)) +
  theme_bw()
```

**Figure `r i=i+1; i`.** Life history parameters, distributions and relationships between them all values are logged.


```{r fig2,eval=TRUE,fig.height=6,fig.width=6}
teleost=transform(teleost,l50linf=l50/linf)

family=transform(teleost,
                 what=ifelse(family%in%c("Scombridae"),
                                  ac(family),"other"))

pc=princomp(teleost[,c("linf","k","b","l50linf")],
            cor=TRUE,use="pairwise.complete.obs")
gg=ggbiplot(pc, obs.scale=1, var.scale=1, 
                ellipse=TRUE, ellipse.prob=.5, circle=FALSE,
         groups=factor(family[,"what"]))+
  kobe:::theme_ms(10,legend.position="bottom")
gg=gg+geom_point(aes(xvar,yvar),data=gg$data[130,],size=3)

#gg$layers[[2]]$mapping$colour=NULL
#gg$layers[[3]]$mapping$colour=NULL
gg+theme(legend.position="none") +
   scale_colour_manual(values=c("red","blue"))#guides(col=guide_legend(nrow = 3))
```

**Figure `r i=i+1; i`.** 

```{r fig3}
source('~/Desktop/flr/git/FLife/R/lhEql.R')
refpts=FLBRP:::refpts

srr=FLBRPs("Beverton and Holt"     =lhEql(alb,sr="bevholt"),
           "Ricker"                =lhEql(alb,sr="ricker"),
           "Cushing"               =lhEql(alb,sr="cushing"),
           "Shepherd"              =lhEql(rbind(alb,FLPar(c=1.5)),sr="shepherd"),
           "Segmented \nRegression"=lhEql(alb,sr="segreg"))

srr=
  ldply(srr,function(x) {
  refpts(x)=refpts(x)["msy"]
  fbar(x)=seq(0,1,length.out=501)
  res=brp(x)
  subset(model.frame(FLQuants(res,"ssb","rec","catch"),drop=TRUE),ssb>=0)})

ggplot(melt(srr[,-5],id=c("year","ssb",".id")))+
  geom_vline(aes(xintercept=200))+
  geom_line(aes(ssb,value,col=.id))+
  theme_bw()+theme(legend.position="bottom")+
  scale_colour_manual("Stock Recruit \n Relationship",
                      values=c("red","green","yellow","blue","pink"))+
  xlab("Spawning Stock Biomass")+ylab("Recruits")
```

**Figure `r i=i+1; i`.** Stock recruitment relationships for a steepness of 0.75 and vigin biomass of 1000.

```{r fig4}
ggplot(melt(srr[,-4],id=c("year","ssb",".id")))+
  geom_path(aes(ssb,value,col=.id))+
  theme_bw()+theme(legend.position="bottom")+
  scale_colour_manual("Stock Recruit \n Relationship",
                      values=c("red","green","yellow","blue","pink"))+
  xlab("Spawning Stock Biomass")+ylab("Yield")
```

**Figure `r i=i+1; i`.** Production curves, Yield v SSB, for a steepness of 0.75 and vigin biomass of 1000.


```{r fig5,eval=TRUE,fig.height=3,fig.width=6}
par=lhPar(teleost[,c("linf","k","t0","l50","a","b")])
attributes(par)$species=teleost[,"species"]
attributes(par)$family =teleost[,"family"]
family=transform(teleost,
                 what=ifelse(family%in%c("Scombridae"),
                                  ac(family),"other"))

scen=expand.grid(juve =!c(TRUE,FALSE),
                 dome =!c(TRUE,FALSE),
                 steep=c(0.75,.9),
                 m    =c("Gislason","Constant"))

species=attributes(par)$species

mGislason=function(length,params) 
   0.55*(length^-1.66)%*%(params["linf"]^1.44)%*%params["k"]
mConstant=function(length,params){
      length[]=params["l50"]
      0.55*(length^-1.66)%*%(params["linf"]^1.44)%*%params["k"]}

ref=mdply(scen,function(dome,juve,steep,m){
  par["s"]  =steep
  par["a1",]=ifelse(juve,.75,1)*par["a1",]
  par["sr",]=ifelse(dome,5,5000)
    
  if (m=="Gislason"){
     res=lhRef(par,m=mGislason)}
  else{
     res=lhRef(par,m=mConstant)}
  
  data.frame(species,res)})
```

**Figure `r i=i+1; i`.** 

```{r}
dat=ref[dimnames(ref[!is.na(ref[,"rc"]),])[[1]],]
           
pc=princomp(dat[,c("r","rc","lopt","sk","lro")],
            cor=TRUE,use="pairwise.complete.obs")

gg=ggbiplot(pc, obs.scale=1, var.scale=1, 
                ellipse=TRUE, ellipse.prob=.5, circle=FALSE,
         groups=factor(rep(family[,"what"],16)))+
  kobe:::theme_ms(12,legend.position="bottom")

gg=gg+geom_point(aes(xvar,yvar),data=gg$data[130,],size=3)

gg$layers[[2]]=NULL
#gg$layers[[2]]$mapping$colour=NULL
#gg$layers[[3]]$mapping$colour=NULL
gg+theme(legend.position="none")+
   scale_colour_manual(values=c("red","blue"))
```

**Figure `r i=i+1; i`.** 

```{r}
#alb["t0"]=-alb["t0"]
## Beverton and Holt recruitment
bh      =lhEql(alb,m=lorenzen)
refpts(bh)=refpts(bh)["msy",]
source('~/Desktop/flr/git/FLBRP/R/plot-ggplot.R')
p=plot(bh)+theme_bw()+theme(legend.position="bottom")+
  scale_colour_manual("",values="red",label="MSY")
```

```{r DD,eval=FALSE}
library(FLife)

source("/home/laurie/Desktop/flr/git/FLife/R/lorenzen.R")

alb=FLPar(unlist(teleost[teleost$species=="Thunnus alalunga",
        c("linf","k","t0","l50","a","b")]))
alb=lhPar(rbind(alb,FLPar(m1=0.15,m2=-0.288,s=0.75)))

## Beverton and Holt recruitment
bh      =lhEql(alb,m=lorenzen)

fbar(bh)=FLQuant(c(seq(0,4,length.out=101)))*
              refpts(bh)["msy","harvest"]
names(dimnames(fbar(bh)))[1]="age"
stk=FLStocks("SRR"=as(bh,"FLStock"))
stk[["SRR"]]=fwd(stk[["SRR"]],f=fbar(bh)[,-1],sr=bh)

## Cushing recruitment
#sr=fmle(as.FLSR(stk[[1]][,1:10],model="geomean"))
cu        =lhEql(alb,m=lorenzen,sr="cushing")
fbar(cu)  =fbar(bh)
names(dimnames(fbar(cu)))[1]="age"
stk[["M"]]        =as(cu,"FLStock")
stk[["M"]]        =fwd(stk[["M"]],f=fbar(cu)[,-1],sr=cu)
stk[["Fecundity"]]=stk[["M"]]

eql=FLBRPs("Beverton Holt"=bh,"Cushing"=cu)
rm(bh,cu)

y  =eql[["Cushing"]]
ref=stock.n(y)[,1]
n  =stock.n(y)[,1]
m  =m(y)

mDD=mdply(data.frame(i=seq(dims(fbar(eql[["Cushing"]]))$year)),
          function(i){
         
  repeat{
      scale=(stock.n(y)[,i]%-%ref)%/%ref
      m(y)[] =mdd(stock.wt(y),alb,scale,k=.9)
    
    if(sum((m-m(y))^2)<1e-6){break}
    m=m(y)}          
    
  subset(model.frame(FLQuants(y,"catch","ssb"),drop=TRUE),
         year==i)})

p=ggplot(mDD)+geom_line(aes(ssb,catch))

y  =eql[["Cushing"]]
ref=stock.n(y)[,1]
n  =stock.n(y)[,1]
mat=mat(y)#[,max(i-1,dims(m(y))$year)]    

matDD=mdply(data.frame(i=seq(dims(fbar(eql[["Cushing"]]))$year)[-1]),
          function(i){
            
  repeat{
      scale=(stock.n(y)[,i]%-%ref)%/%ref
      mat(y)[] =matdd(ages(scale),alb,scale,k=.9)
  
    if(sum((mat-mat(y))^2)<1e-6){break}
    mat=mat(y)}
  
  subset(model.frame(FLQuants(y,"catch","ssb"),drop=TRUE),
         year==i)})
```


```{r fig6,fig.height=5,fig.width=6,eval=FALSE}
data=rbind.fill(
     cbind("Form"="M",mDD),cbind("Form"="Fecundity",matDD),
     cbind("Form"="Stock-recruit",model.frame(FLQuants(eql[["Cushing"]],"catch","ssb"),drop=TRUE)))

ggplot(data)+
   geom_line(aes(ssb,catch,colour=Form))+theme_bw()+theme(legend.position="bottom")+
  scale_colour_manual("",values=c("red","green","blue"))+
  xlab("Biomass")+ylab("Yield")
```

```{r fig7,fig.height=6}
source('~/Desktop/flr/git/FLife/R/mat-dd.R')

k   =1
sr  ="cushing"
s   =0.7
alb  =FLPar(unlist(teleost[teleost$species=="Thunnus alalunga",
             c("linf","k","t0","l50","a","b")]))
alb  =lhPar(rbind(alb,FLPar(m1=0.3,m2=-0.288,s=s)))
eq      =lhEql(alb,m=lorenzen,sr=sr)
fbar(eq)=FLQuant(c(seq(0,5,length.out=51)*
                 refpts(eq)["msy","harvest"]))
f       =fbar(eq)[,-1]

stkr=as(eq,"FLStock")
stkr=fwd(stkr,f=f,sr=eq)
ref =stock.n(eq)[,1]

stkm=stkr
m(  stkm)=mdd(stock.wt(stkm),alb,(stock.n(stkm)%-%ref)%/%ref,k) 
stkm=fwd(stkm,f=f,sr=eq)

stkf=stkr
mat(stkf)=matdd(ages(stock.wt(stkf)),alb,
                (stock.n(stkf)%-%ref)%/%ref,k,TRUE) 
stkf=fwd(stkf,f=f,sr=eq)

for (i in seq(dims(fbar(eq))$year)[2:50]){
  scale =(stock.n(stkm)[,i]%-%ref)%/%ref
  m(stkm)[,i]=mdd(stock.wt(stkm)[,i],alb,scale,k)
            
  stkm=fwd(stkm,f=f[,i],sr=eq)}

for (i in seq(dims(fbar(eq))$year)[2:50]){
  scale =(stock.n(stkf)[,i]%-%ref)%/%ref
  mat(stkf)[,i]=matdd(ages(scale),alb,scale,k,TRUE)
            
  stkf=fwd(stkf,f=f[,i],sr=eq)}

plot(FLStocks("SRR"=stkr[,-(1:1)],
              "M"  =stkm[,-(1:1)],
              "Fecundity"=stkf[,-(1:1)]))+
  theme_bw()+
  theme(legend.position="bottom")

p=ggplot(as.data.frame(FLQuants("SRR"=ssb(stkr),
                              "M"=ssb(stkm),
                              "Fecundity"=ssb(stkf))))+
  geom_line(aes(year,data,col=qname))+
  theme_bw()+
  theme(legend.position="bottom")
```


```{r fig8,fig.height=6}
k   =1
sr  ="cushing"
s   =0.7
alb  =FLPar(unlist(teleost[teleost$species=="Thunnus alalunga",
             c("linf","k","t0","l50","a","b")]))
alb  =lhPar(rbind(alb,FLPar(m1=0.3,m2=-0.288,s=s)))
eq      =lhEql(alb,m=lorenzen,sr=sr)
fbar(eq)=FLQuant(c(seq(5,.5,length.out=51)*
                 refpts(eq)["msy","harvest"]))
f       =fbar(eq)[,-1]

stkr=as(eq,"FLStock")
stkr=fwd(stkr,f=f,sr=eq)
ref =stock.n(eq)[,1]

stkm=stkr
m(  stkm)=mdd(stock.wt(stkm),alb,(stock.n(stkm)%-%ref)%/%ref,k) 
stkm=fwd(stkm,f=f,sr=eq)

stkf=stkr
mat(stkf)=matdd(ages(stock.wt(stkf)),alb,
                (stock.n(stkf)%-%ref)%/%ref,k,TRUE) 
stkf=fwd(stkf,f=f,sr=eq)

for (i in seq(dims(fbar(eq))$year)[2:50]){
  scale =(stock.n(stkm)[,i]%-%ref)%/%ref
  m(stkm)[,i]=mdd(stock.wt(stkm)[,i],alb,scale,k)
            
  stkm=fwd(stkm,f=f[,i],sr=eq)}

for (i in seq(dims(fbar(eq))$year)[2:50]){
  scale =(stock.n(stkf)[,i]%-%ref)%/%ref
  mat(stkf)[,i]=matdd(ages(scale),alb,scale,k,TRUE)
            
  stkf=fwd(stkf,f=f[,i],sr=eq)}

plot(FLStocks("SRR"=stkr[,-(1:1)],
              "M"  =stkm[,-(1:1)],
              "Fecundity"=stkf[,-(1:1)]))+
  theme_bw()+
  theme(legend.position="bottom")

p=ggplot(as.data.frame(FLQuants("SRR"=ssb(stkr),
                              "M"=ssb(stkm),
                              "Fecundity"=ssb(stkf))))+
  geom_line(aes(year,data,col=qname),size=2)+
  theme(legend.position="bottom")
```

```{r fig9}
data(ple4)
res=noise(4,m(ple4)[1:8,ac(1980:2008)],burn=10,b=0.9,what="age")
ggplot()+
  geom_point(aes(year,age,size= data),
             data=subset(as.data.frame(res),data>0))+
  geom_point(aes(year,age,size=-data),
             data=subset(as.data.frame(res),data<=0),colour="red")+
  scale_size_area(max_size=4, guide="none")+
  facet_wrap(~iter)+theme_bw()
```

```{r fig10}
source('~/Desktop/flr/git/FLife/R/noise.R')
res=noise(4,m(ple4)[1:8,ac(1980:2008)],burn=10,b=0.9,what="cohort")
ggplot()+
  geom_point(aes(year,age,size= data),
             data=subset(as.data.frame(res),
                         data>0))+
  geom_point(aes(year,age,size=-data),
             data=subset(as.data.frame(res),
                         data<=0),colour="red")+
  scale_size_area(max_size=4,  guide="none")+
  facet_wrap(~iter)+theme_bw()  
```

```{r fig11, fig.height=6,fig.width=10}
k   =1
sr  ="cushing"
s   =0.7
alb  =FLPar(unlist(teleost[teleost$species=="Thunnus alalunga",
             c("linf","k","t0","l50","a","b")]))
alb  =lhPar(rbind(alb,FLPar(m1=0.3,m2=-0.288,s=s)))
eq      =lhEql(alb,m=lorenzen,sr=sr)
fbar(eq)=FLQuant(rep(1,1001)*refpts(eq)["msy","harvest"])
f       =fbar(eq)[,-1]
f       =propagate(fbar(eq)[,-1],3)
f[,,,,,2]=0
f[,,,,,3]=f[,,,,,1]*4

stkr=as(eq,"FLStock")
stkr=fwd(stkr,f=f,sr=eq)
ref =stock.n(eq)[,1]

scale=noise(1,stock.wt(stkr),sd=0.3,b=0.9,what="cohort")
stkm=stkr

m(  stkm)=mdd(iter(stock.wt(stkm),1),alb,scale,k) 
stkm=fwd(stkm,f=f,sr=eq)

scale=noise(1,stock.wt(stkm),sd=0.3,b=0.9,what="cohort")
stkf=stkr
mat(stkf)=matdd(ages(stock.wt(stkf)),alb,scale,k,TRUE) 
stkf=fwd(stkf,f=f,sr=eq)
stkr=fwd(stkr,f=f,sr=eq,sr.residuals=rlnorm(1,iter(f,1)*0,.3))
dat =as.data.frame(FLQuants("SRR"=ssb(stkr),
                              "M"=ssb(stkm),
                              "Fecundity"=ssb(stkf)))
fishMat=ddply(subset(dat,year>50),.(iter,qname), transform, val=data/mean(data))

k   =1
sr  ="cushing"
s   =0.7
alb  =FLPar(unlist(teleost[teleost$species=="Thunnus alalunga",
             c("linf","k","t0","l50","a","b")]))
alb  =lhPar(rbind(alb,FLPar(m1=0.3,m2=-0.288,s=s)))
alb[c("a1","sr")]=c(0,5000)
eq      =lhEql(alb,m=lorenzen,sr=sr)
fbar(eq)=FLQuant(rep(1,1001)*refpts(eq)["msy","harvest"])
f       =fbar(eq)[,-1]
f       =propagate(fbar(eq)[,-1],3)
f[,,,,,2]=0
f[,,,,,3]=f[,,,,,1]*4

stkr=as(eq,"FLStock")
stkr=fwd(stkr,f=f,sr=eq)
ref =stock.n(eq)[,1]

scale=noise(1,stock.wt(stkr),sd=0.3,b=0.9,what="cohort")

stkm=stkr
m(  stkm)=mdd(iter(stock.wt(stkm),1),alb,scale,k) 
stkm=fwd(stkm,f=f,sr=eq)
scale=noise(1,stock.wt(stkm),sd=0.3,b=0.9,what="cohort")
stkf=stkr
mat(stkf)=matdd(ages(stock.wt(stkf)),alb,scale,k,TRUE) 
stkf=fwd(stkf,f=f,sr=eq)
stkr=fwd(stkr,f=f,sr=eq,sr.residuals=rlnorm(1,iter(f,1)*0,.3))
dat =as.data.frame(FLQuants("SRR"=ssb(stkr),
                            "M"=ssb(stkm),
                            "Fecundity"=ssb(stkf)))
fishJuv=ddply(subset(dat,year>50),.(iter,qname), transform, val=data/mean(data))

dat=rbind(cbind("Selection"="Juvenile",fishJuv),
          cbind("Selection"="Mature",  fishMat))
dat=transform(dat,F=factor(paste("F times",c(0,1,3))[iter]))

ggplot(dat)+
  geom_line(aes(year,val,col=qname))+
  theme_bw()+
  theme(legend.position="bottom")+
  facet_grid(F~Selection,scale="free_y")+
  scale_x_continuous(limits=c(500,700))+
  xlab("Time (year)")+ylab("")
```


```{r fig12,fig.width=6,fig.height=6}
dat=rbind(cbind("Selection"="Juvenile",fishJuv),
          cbind("Selection"="Mature",  fishMat))
dat=transform(dat,F=factor(paste("F times",c(0,1,3))[iter]))

dat=ddply(dat,.(Selection,qname,F), with, 
#          as.data.frame(spectrum(data, spans = c(7,13), log = "dB", ci = 0.8,plot=FALSE)[c("freq","spec")]))
          as.data.frame(spectrum(data, log = "dB", ci = 0.8,plot=FALSE)[c("freq","spec")]))

dat=ddply(subset(dat,freq>0.05),.(F,Selection,qname),transform,val=spec/max(spec))
ggplot(dat,aes(freq,val,col=qname))+
  geom_smooth(se=FALSE)+
  facet_grid(F~Selection,scale="free_y")+
  theme_bw()+
  theme(legend.position="bottom")
```


```{r fig13, fig.width=10,fig.height=4}
library(FLAssess)

stk   =iter(stkm[,101:250],1)
m(stk)=iter(m(stkr),1)[,101:250]
vpa    =stk+VPA(stk)

p=plot(FLStocks("Actual"=stk,"Model Estimate"=vpa))
p$data=subset(p$data,qname%in%c("SSB","Rec"))
names(p$data)[8]="data"
p$data=ddply(p$data,.(qname), transform, val=data/mean(data))
ggplot(subset(p$data,year>50))+
  geom_line(aes(year,val,col=qname))+
  facet_grid(qname~stock,scale="free")+
  theme_bw()+
  theme(legend.position="none")+
  xlab("Year")+ylab("")
```


```{r fig14, fig.width=6,fig.height=8}
library(FLife)
library(FLash)
library(FLBRP)
  
par=lhPar(FLPar(linf=100))  
eql=lhEql(par)
mou=as(eql,"FLStock")
mou=FLash:::fwd(mou,f=fbar(eql)[,-1]/4,sr=eql)
  
srDev=rlnorm(100,FLQuant(rep(0,136)),0.3)
  
om=FLife:::mseSBT1(mou,eql,srDev,
             start=dims(mou)$maxyear,end=dims(mou)$maxyear+20,lag=1,interval=3,
             k1=1.5,k2=3.0,gamma=1,nyrs=5,   
             seed=7890,nits=100,
             uCV =0.2) 
  
ggplot()+
    geom_line(aes(year,data,col=iter),
              data=as.data.frame(FLQuants(iter(om[,ac(95:120)],5:7),c("Rec"=rec,"SSB"=ssb,
                                                      "Catch"=catch,"Harvest"=fbar)),drop=T))+
    facet_grid(qname~.,scale="free")+
    theme_bw()+xlab("")+ylab("")+
  theme(legend.position="bottom")
```


```{r fig15,fig.height=10}
cas=read.csv("/home/laurie/MEGAsync/mse/trade-offs/inputs/casSWOM8513_v1.csv",
             colClasses=c("factor","numeric","factor","factor","factor","factor","numeric","numeric","numeric"))
names(cas)=c("species","year","flag","fleet","gear","stock","len","ln5","n")
powh=function(len,n){
  
  fn=function(len,n){
    require(plyr)
    
    res=ddply(data.frame(n=n,len=len), .(len), function(x) sum(x$n))
    res=res[order(res$len),]
    
    csum =rev(cumsum(rev(res$V1)))
    clsum=rev(cumsum(rev(res$len*res$V1)))
    mn   =clsum/csum
    
    data.frame(mn=mn,diff=mn-res$len,len=res$len,n=res$V1)}
  
  linf=function(x) -coefficients(x)[1]/coefficients(x)[2]
  zk  =function(x) (-1-coefficients(x)[2])/coefficients(x)[2]
 
  dat=fn(len,n)
  
  res=lm(diff~len,data=dat)
  
  params=c("linf"=linf(res),"zk"=zk(  res))
  names(params)=c("linf","zk")
  
  return(list(params=params,
              data  =dat))}

pw=ddply(subset(cas), .(year), 
            function(cas) powh(cas$len,cas$n)$data)
      
pw=transform(pw, lustrum=(year%/%5)*5,
                 yr    =year-(year%/%5)*5,
                 weight=ifelse(len>=100&len<=200,1,0))

ggplot(pw)+
  geom_line(aes(len,diff,colour=factor(yr),group=year))+
  scale_x_continuous(limits=c(0,300)) +
  facet_wrap(~lustrum,ncol=2)+
  xlab("Length (cm)")+
  ylab("Difference between Length and Mean Size")+
  geom_smooth(aes(len,diff,weight=weight),
              method="lm",col="red",size=1.25,alpha=.1)+
  theme_bw()+theme(legend.position="none")
```
